---
title: "10.8.2026Kratfrequencydistributions"
format: html
editor: visual
---

## Set up

```{r}
library(ggplot2)
library(brms)
library(ggeffects)
library(tidyverse)
library(readxl)
library(here)
library(dplyr)
library(ggtext)

#Load in clean data set
df.hlo <- read.csv("10.8.2026KratCleanDataRMC.csv") #w/o head length outlier 36 mm
head(df)

df <- read.csv("RC2026_KRAT_DATA.08.01.csv")
```

# Goal

Create frequency distributions for general sizes of each species.

Create frequency distributions for each species over the different seasons of the year.

## Our data demographics

```{r}
#Remove rows with the most NAs or where most data was not taken
df <- df[-c(6, 10, 11, 12,13, 14,15, 16, 19, 20, 21, 22, 23, 24, 25, 34, 36, 37, 38, 39, 51, 57, 60, 61, 66, 69, 70, 76, 78, 79, 80, 86, 90, 91, 92, 93, 95, 96), ]

#Show number of each species included in analysis: 36 heermanni, 24 venustus
table(df$Species)
```

## General size frequency distributions

#### With all data points

```{r}
#Create a histogram of the head length, separating by species
df %>%
  ggplot(aes(x= Head_Length, fill =  Species))+
geom_histogram(bins = 20, color = "black")

#Create a density, smooth line frequency distribution, separating by species
df %>%
  ggplot(aes(x= Head_Length, color =  Species))+
geom_density()
```

#### Without head length outlier (\< 37 mm)

```{r}
#Create a histogram of the head length, separating by species
df.hlo %>%
  ggplot(aes(x= Head_Length, fill =  Species))+
geom_histogram(bins = 20, color = "black")

#Create a density, smooth line frequency distribution, separating by species
df.hlo %>%
  ggplot(aes(x= Head_Length, color =  Species))+
geom_density()
```

Tail is left out of the trait distribution when the head length value \<37 mm is not included. Does this mean its actually an outlier?

#### Spring data showing the outlier

The more data collected over the summer filled in the gaps of the distribution that had been missing in this spring data analysis. This head length value \< 37 may not be an outlier after all??

```{r}
#Read in old data from the Spring, May 2026
df.old <- read.csv("final_krat_data_2026.csv")

#Remove rows where most data was not taken (recaps mostly)
df.old <- df.old[-c(6, 10, 11, 12,13, 14,15, 16, 19, 20, 21, 22, 23, 24, 25, 34, 36, 37, 38, 39, 51, 57, 60, 61, 66), ]

#Have 27 heermanni, 14 venustus in this analysis from the spring before summer data was taken + included
table(df.old$Species)

#Create a density, smooth line frequency distribution, separating by species
df.old %>%
  ggplot(aes(x= Head_Length, color =  Species))+
geom_density()
```

# Size over the seasons

```{r}
#make a new column for season
#sort observations by date -- choose what counts as each season
```
