---
title: "10.08.2026: Kangaroo rat Measurement outliers and graphs"
format: html
editor: visual
---

## Set up

```{r}
library(ggplot2)
library(brms)
library(ggeffects)
library(tidyverse)
library(readxl)
library(here)
library(dplyr)
library(ggtext)

#Load in clean data set
df <- read.csv("RC2026_KRAT_DATA.08.01.csv")
head(df)
```

# Goal:

Graph each continuous quantitative variable versus head length. Get rid of outliers. Decide which variables are redundant and do not change much with species. These can be removed from the PCA later.

## 1. Assess head length for any outliers

```{r}
#In this plot we can see a head_length value of about 36 which could be an outlier??
df %>%
  ggplot(aes(y= Head_Length, x= Body_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
#Plotting versus tail length, we can see it is far out here as well
df %>%
  ggplot(aes(y= Head_Length, x= Tail_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
```

### Remove head length outlier

```{r}
#Create new copy data frame
df1 <- df
#Remove possible outlier
df1$Head_Length[df$Head_Length < 37] <- NA

#Write new csv with the new change
#write_csv(df1, "10.8.2026KratCleanDataRMC.csv")

#Load in clean data set without head length outlier (hlo)
df.hlo <- read.csv("10.8.2026KratCleanDataRMC.csv") #w/o head length outlier 36 m
```

# 2. Head length outlier may not be an outlier

## Our data demographics

### All up to date, data points from Summer 2026

36 heermanni, 24 venustus

```{r}
#Remove rows with the most NAs in continuous, quantitative variables or where most measurement data was not taken (Mostly recaps)
df <- df[-c(6, 10, 11, 12,13, 14,15, 16, 19, 20, 21, 22, 23, 24, 25, 34, 36, 37, 38, 39, 51, 57, 60, 61, 66, 69, 70, 76, 78, 79, 80, 86, 90, 91, 92, 93, 95, 96), ]  #Show number of each species included in analysis: 36 heermanni, 24 venustus 
table(df$Species)
```

## General size frequency distributions for Head Length

#### With all data points

Option here between a histogram and a density distribution. I like the density one better for looks, but probably tells you less if you don't know about the data for sample size, species counts and the actual amount the density assesses.

```{r}
#Create a histogram of the head length, separating by species 
df %>%   ggplot(aes(x= Head_Length, fill =  Species))+ geom_histogram(bins = 20, color = "black")  
#Create a density, smooth line frequency distribution, separating by species 
df %>%   ggplot(aes(x= Head_Length, color =  Species))+ geom_density()
```

#### Without head length outlier (\< 37 mm)

```{r}
#Create a histogram of the head length, separating by species without the outlier
df.hlo %>%   ggplot(aes(x= Head_Length, fill =  Species))+ geom_histogram(bins = 20, color = "black")  #Create a density, smooth line frequency distribution, separating by species without the outlier
df.hlo %>%   ggplot(aes(x= Head_Length, color =  Species))+ geom_density()
```

Tail is left out of the trait distribution when the head length value \<37 mm is not included. Does this mean its actually an outlier?

#### Old spring data showing the outlier (what we looked at in meeting 10/8/26)

The more data collected over the summer filled in the gaps of the distribution that had been missing in this spring data analysis. This head length value \<37 mm may not be an outlier after all??

Observation demographics of old data: 27 heermanni, 14 venustus

```{r}
#Read in old data from the Spring, May 2026
df.old <- read.csv("final_krat_data_2026.csv")

#Remove rows where most data was not taken (recaps mostly)
df.old <- df.old[-c(6, 10, 11, 12,13, 14,15, 16, 19, 20, 21, 22, 23, 24, 25, 34, 36, 37, 38, 39, 51, 57, 60, 61, 66), ]

#Have 27 heermanni, 14 venustus in this analysis from the spring before summer data was taken + included
table(df.old$Species)

#Create a density, smooth line frequency distribution, separating by species
df.old %>%
  ggplot(aes(x= Head_Length, color =  Species))+
geom_density()
```

This might be explained by more juveniles being trapped in the early summer. If you look at the seasonal frequency distributions (in other analysis) this shows some pretty good evidence of size differences by seasons in Dipodomys heermanni.

# 3. Why head length as a size proxy?

#### Why head length and not body length as a size proxy?

Head length is being used due to its value as a general size indicator and also because it is a less coarse measurement than body length. When measuring small mammals a different scruff can change the body length and it is really a general estimate. The head length was taken with a caliper and is highly precise.

Additionally, we will see this later in data, body length is almost entirely overlapping between D. heermanni and D. venustus from our sample meaning body length would not help to differentiate species in a PCA analysis and head length would be a better size proxy to help find separation between the two.

### Explore body length vs. head length as a measure of general size

#### Frequency distributions of Head length

```{r}
#Create a density, smooth line frequency distribution, separating by species
df %>%
  ggplot(aes(x= Head_Length, color =  Species))+
geom_density()
```

#### Frequency distributions of body length

```{r}
#Create a density, smooth line frequency distribution of body length, separating by species 
df %>%
  ggplot(aes(x= Body_Length, color =  Species))+
geom_density()
```

#### Interpretation:

Body length seems to be very similar and nearly entirely overlapping on both species, but head length is different, but overlapping. This could mean that the ratio of head to body length is different on these animals.

#### Ratio of body length to head length versus body length

```{r}
#Show data of smooth linear regression plot
df |>
  ggplot(aes(y= log(Body_Length/Head_Length), x= log(Body_Length), color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Show body length to head length ratio in a density frequency distribution plot
df %>%
  ggplot(aes(x= (Body_Length/Head_Length), color =  Species))+
geom_density()
```

They don't seem to be too different and the body length to head length ratio would not help us to identify between these species very accurately. The head length on its own is more helpful than including body length. Consider excluding body length from PCA?

# 4. Graph each measurement versus head length

Only quantitative, continuous measurements will be graphed for outliers

```{r}
#Here are all of the variables measured
colnames(df)
```

Quantitative, continuous measurements to try:

#### 1) Body length

I'm trying logged versus normal and it doesn't seem to make a noticeable difference in the graphs. It only seems to get rid of real biological references like the millimeter units from the graph making it harder to understand.

```{r}
#Body length vs head length as logged
df %>%
  ggplot(aes(y= log(Body_Length), x= log(Head_Length), color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
#body length vs head length, not logged
df %>%
  ggplot(aes(y= Body_Length, x= Head_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
```

### 2) Tail length

```{r}
#Tail length vs head length graph
df %>%
  ggplot(aes(y= Tail_Length, x= Head_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Tail length vs head length
df %>%
  ggplot(aes(y= Tail_Length, x= Body_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
```

#### Outliers:

There seem to be 2 outliers in the D. venustus with tail lengths \< 200 mm. When trapping its very possible that those kangaroo rats had the tip of its chopped or bitten off shortening it. I actually do remember some individuals like that at Hastings.

I looked back at these two individuals with tail lengths of \~178 mm, they were both trapped on the same day 6/22/2025 which early summer when we expect to see grown up juvenile-ish young adult kangaroo rats of the season. One of these individuals, tag 108 was captured again on 11/08/2025 and had a longer tail of 211 mm. This proves that these outliers really just represent younger individuals that are still growing.

This could be the same case for the D. heermanni with a tail length \< 160 mm. Looking back in the notes of the data set the individual, tag F002 trapped on 6/30/2026 is a juvenile, proving that these outliers are due to lifestage and represent important seasonal and lifestage variation.

#### Interpretation:

This data hints that young adult kangaroo rats of the season may not differ greatly from other adults in head length or body length, but do differ greatly in tail length. This means tail length could have the potential to be a good indicator of age and life-stage in kangaroo rats and picking out the young of the year in data sets.

### 3) Weight (animal only)

Side note: When I was working on getting rid of NAs for weight values, for some values we had a bag + animal weight but no bag weight, and I could estimate the bag weight for about 5 values to fill up this data. TBD

```{r}
#Weight of animal versus head length graph
df %>%
  ggplot(aes(y= Weight_Animal_Only, x= Head_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
```

#### Outliers:

There seems to be 1 really crazy outlier for the animal weight where it is \< 30 grams which is insane. This was D. venustus individual, tag 843 that has a animal only weight of 15 grams. This is a data taker error, as there are other data taker errors in this observation. This value of 15 grams is pretty much how much the bags weigh and the bag+animal weight is shown to 115 grams, so I can be very sure this animal is actually 100 grams not 15g.

(This becomes clear once this was graphed again later, but there is another possible outlier for D. venustus that is less \<70 g. Looking through notes, this D. venustus individual, tag B135, trapped 6/23/2026 has a weight of 64 g and is a young adult female. This difference can largely be explained by a different lifestage in our data. (interestingly the tail is really long 222 mm disproving the last point of short tail = young, atleast in this case)

#### Fix outliers:

```{r}
#Create new dataframe df.o (dataframe.outliers) which will include all my edits to remove outliers
df.o <- df
#Fix outlier to true value, explained why above
df.o$Weight_Animal_Only[df$Weight_Animal_Only == 15] <- 100

#Plot without the outliers
df.o %>%
  ggplot(aes(y= Weight_Animal_Only, x= Head_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
```

#### Interpretation:

This data shows that animal weight and head length together are very good indicators of species between these two!! With some weights and head lengths that overlap and can be tricky. Around 75 - 85 g weight and 42 - 46 mm head length are the overlaps which species cannot be told here.

### 4) Ear length

```{r}
#graph the ear length vs head length
df %>%
  ggplot(aes(y= Ear_Length, x= Head_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
```

#### Outliers:

Theres a crazy outlier for ear length that is 125 mm, which is likely explained as a typing error. That is likely a tail length or body length.

#### Fix outliers

```{r}
#Exclude any ear length values greater than 25 mm (already a crazy number for ear length here)
df.o$Ear_Length[df$Ear_Length > 25] <- NA

#Regraph with outliers removed
df.o %>%
  ggplot(aes(y= Ear_Length, x= Head_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
```

#### Interpretation:

The ear length and head length seem to be pretty good at distinguishing between species for most values. Between 16-18 mm of ear length and 42-46 mm head length we cannot be sure of species.

### 5) Foot length

```{r}
#Graph foot length vs head length
df %>%
  ggplot(aes(y= Foot_Length, x= Head_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
```

#### Outliers:

I'm not really sure with this graph since more than half of the values are outside of the confidence intervals and the data is consistently super spread.

Looking at notes for D. heermanni individual with footlength of 48 mm. Individual was tag #12 caught in 7/12/2025, and this is the same individual which had a data entry error for the ear of 125 mm (crazy insane totally data taker fault), this could suggest the data taker potential may not have typed this correctly??? But it doesn't seem to be that crazy for a kangaroo rat foot length or this species if looking into the literature.

#### Interpretation:

There are clear differences in mean foot length between these species, but this trait seems to be incredibly variable with a lot of overlap among them which would make it hard to distinguish based on foot length when looking at a single individual.

### 6)  Foot width

```{r}
#Graph foot width vs head length
df %>%
  ggplot(aes(y= Foot_Width, x= Head_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
```

### Outliers + Interpretation:

D. venustus individual with \~6.5 mm foot width is pretty far out. Tag 843 has a foot width value of 6.25 mm when it was caught 4/11/2026 and then it has a lower foot width value of 5.62 mm when it was caught about 2.5 months later on 6/23/2026. It seems like this value should not decrease as the animal ages and could either suggest that this measurement is no good, or that maybe the muscle mass increases/ decreases over months??? (unlikely?)

### 7) Hind limb length

```{r}
#Graph hind limb length vs head length
df %>%
  ggplot(aes(y= Hind_Limb_Length, x= Head_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
```

### Outliers:

The data seems pretty spread

8\) Toe length

9\) Forelimb length

10\) Finger length

11\) Wrist width

12\) Hand width

13\) Head length

14\) Maxillary head width

15\) Bolea head width
