---
title: "10.8.2026: Kangaroo rat frequency distributions over size, sex and seasons"
format: pdf
editor: visual
---

## Set up

```{r}
library(ggplot2)
library(brms)
library(ggeffects)
library(tidyverse)
library(readxl)
library(here)
library(dplyr)
library(ggtext)

#Load in clean data set with all values
df <- read.csv("RC2026_KRAT_DATA.08.01.csv")
```

# Goal

Create frequency distributions for general sizes of each species.

Create frequency distributions for each species over the different seasons of the year.

## Our data demographics

36 heermanni, 24 venustus

```{r}
#Remove rows with the most NAs in continuous, quantitative variables or where most measurement data was not taken
df <- df[-c(6, 10, 11, 12,13, 14,15, 16, 19, 20, 21, 22, 23, 24, 25, 34, 36, 37, 38, 39, 51, 57, 60, 61, 66, 69, 70, 76, 78, 79, 80, 86, 90, 91, 92, 93, 95, 96), ]

#Show number of each species included in analysis: 36 heermanni, 24 venustus
table(df$Species)
```

## General size frequency distributions for Head Length

#### With all data points as of 8.01.2026

Option here between a histogram and a density distribution. I like the density one better for looks, but probably tells you less if you don't know about the data for counts and the actual frequency.

```{r}
#Create a histogram of the head length, separating by species
df %>%
  ggplot(aes(x= Head_Length, fill =  Species))+
geom_histogram(bins = 20, color = "black")

#Create a density, smooth line frequency distribution, separating by species
df %>%
  ggplot(aes(x= Head_Length, color =  Species))+
geom_density()
```

## General size frequency distribution for Body Length

#### With all data points

This graph has the frequency distribution for body length of both species, showing they are not very different in body length, but in the past graphs the head length has differed a lot more. Maybe body length is a better indicator of general size despite it being more coarse??

```{r}
#Create a density, smooth line frequency distribution, separating by species
df %>%
  ggplot(aes(x= Body_Length, color =  Species))+
geom_density()
```

# Size over the seasons

Observations of both species per season:

Fall: 23, Spring: 3, Summer 33, Winter 1

### Both species across the seasons

```{r}
#make a new column for season
df$Season <- NA
head(df$Season) # it worked!
#sort observations by date -- choose what counts as each season
head(df$Date)

# Extract month and assign season
df <- df %>%
  mutate(
    Month = month(Date),
    Season = case_when(
      Month %in% c(12, 1, 2) ~ "Winter",
      Month %in% c(3, 4, 5)  ~ "Spring",
      Month %in% c(6, 7, 8)  ~ "Summer",
      Month %in% c(9, 10, 11) ~ "Fall"
    )
  )
head(df$Season) # check that it worked
table(df$Season) #Fall 23, Spring 3, Summer 33, Winter 1

#graph with color = season, of all kangaroo rats both species
#Create a density, smooth line frequency distribution, separating by season and including both species taking into account the frequency of body length values
df %>%
  ggplot(aes(x= Body_Length, color =Season))+
geom_density()
```

### Heermanni across the seasons

For just heermanni observations per season:

Fall: 14

Summer: 22

Winter: 0

Spring: 0

```{r}
#Filter for only D. heermanni
head(df$Species)
dfhe <- df |> filter(Species == "Dipodomys heermanni")

#Demographics 14 fall, 22 summer
table(dfhe$Season)

#Create a density, smooth line frequency distribution, separating by season
dfhe %>%
  ggplot(aes(x= Body_Length, color =Season))+
geom_density()
```

This seasonal graph of the heermanni is super interesting and might have enough of a sample size to be able to say something?!

### Venustus over the seasons

For just venustus observations per season:

Fall: 9

Spring: 3

Summer: 11

Winter: 1

```{r}
#Filter for only D. heermanni
head(df$Species)
dfve <- df |> filter(Species == "Dipodomys venustus")

#Demographics 14 fall, 22 summer
table(dfve$Season)

#Create a density, smooth line frequency distribution, separating by season
dfve %>%
  ggplot(aes(x= Body_Length, color =Season))+
geom_density()
```

This graph here is not based on enough data to mean much, but we don't see the same split in size over the seasons as with the heermanni. This could be due mostly to the low sample size.

# Seasonal sexual dimorphism?

### Dipodomys heermanni

Observed 28 Female and 5 Male heermanni.

Females seem to be more abundant then males in both species or maybe they are out foraging more often due to greater metabolic needs (Lactating, pregnant, etc) and therefore get caught in the traps more often??

```{r}
#Create a density, smooth line frequency distribution, of just the heermanni separating by sex and looking at body length frequency
dfhe %>%
  ggplot(aes(x= Body_Length, color =Sex))+
geom_density()

#Sexual demographics
table(dfhe$Sex)
```

This map probably doesn't mean much due to the low sample size of males.

```{r}
#Create a bar plot, separating by season and sex
dfhe %>%
  ggplot(aes(x= Sex, fill=Season))+
geom_bar()
```

There's way more females than males, I'm not sure we can say much from this.

### Dipodomys venustus

Of the venustus observations:

F: 19, M: 1, Unknown: 1

Graph doesn't even work :C

```{r}
#Create a density, smooth line frequency distribution, separating by sex of the venustus and looking at body length value frequency
dfve %>%
  ggplot(aes(x= Body_Length, color =Sex))+
geom_density()

#Sexual demographics
table(dfve$Sex)
```

Graph doesn't even work :C
