---
title: "10.08.2026KratGraphGen"
format: html
editor: visual
---

## Set up

```{r}
library(ggplot2)
library(brms)
library(ggeffects)
library(tidyverse)
library(readxl)
library(here)
library(dplyr)
library(ggtext)

#Load in clean data set
df <- read.csv("RC2026_KRAT_DATA.08.01.csv")
head(df)
```

# Goal:

Graph each continuous quantitative variable versus head length. Decide which variables are redundant and do not change much with species. These can be removed from the PCA later.

#### Why head length and not body length as a size proxy?

Head length is being used due to its value as a general size indicator and also because it is a less coarse measurement than body length. When measuring small mammals a different scruff can change the body length and it is really a general estimate. The head length was taken with a caliper and is highly precise.

## Assess head length for any outliers

```{r}
#In this plot we can see a head_length value of about 36 which seems to be an outlier
df %>%
  ggplot(aes(y= Head_Length, x= Body_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
#Plotting versus tail length, we can see it is an outlier here as well
df %>%
  ggplot(aes(y= Head_Length, x= Tail_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
```

### Remove head length outlier

```{r}
df$Head_Length[df$Head_Length < 37] <- NA
print(df$Head_Length)

write_csv(df, "10.8.2026KratCleanDataRMC.csv")
```
