---
title: "Generating Graphs EEB symposium"
format: html
editor: visual
---

# 06-01-2026 EEB symposium graph generation

## Set up

I will first load the required packages.

```{r}
library(ggplot2)
library(brms)
library(ggeffects)
library(tidyverse)
library(readxl)
library(here)
library(dplyr)
#install.packages("ggtext")
library(ggtext)
krat <- read.csv(here("final_krat_data_2026.csv"))

```

# Exploring Head vs. Body as size proxy

```{r}
head(krat)
#Plot the body length vs head size
krat %>%
  ggplot(aes(y= Head_Length, x = Body_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
#Plotting standard measurements against eachother
krat %>%
  ggplot(aes(x = Head_Length, y = Foot_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
#Plotting body vs foot
krat %>%
  ggplot(aes(x = Body_Length, y = Foot_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
#plotting head length vs ear length
krat %>%
  ggplot(aes(x = Head_Length, y = Ear_Length_Notch, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
#Plotting body length vs ear length
krat %>%
  ggplot(aes(x = Body_Length, y = Ear_Length_Notch, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")+
   coord_cartesian(ylim = c(10, 30)) #crop out outlier (120 mm in ear length)
#Plotting finger length measurements vs Head length measurements
krat %>%
  ggplot(aes(x = Head_Length, y = Finger_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")


```

# Fixing Ear length outlier

I found an outlier with a value over 120 in the ear length row. Here I fixed that issue using the filter fnct. We can see in these graphs the head length differentiates between the two species better than body length, but that no matter what there is significant overlap in measurements when using the ear length

```{r}
#Filter out the ear outlier
kratf <- krat |> filter(Ear_Length_Notch <30)

#plotting head length vs ear length
kratf %>%
  ggplot(aes(x = Head_Length, y = Ear_Length_Notch, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
#Plotting body length vs ear length
kratf %>%
  ggplot(aes(x = Body_Length, y = Ear_Length_Notch, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

```

# Testing graphs

```{r}
#plotting finger length vs ear length
kratf %>%
  ggplot(aes(x = Finger_Length, y = Ear_Length_Notch, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Running bernouilli model to predict species based on ear, finger, head
m.bern.finger.ear.h <- brm(Species ~ 1 + Ear_Length_Notch + Finger_Length+ Head_Length, 
                        data = kratf, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.body") 

#summary model
summary(m.bern.finger.ear.h)
plot(m.bern.finger.ear.h)

#Predict the response based on the model and plot it
pred.m.finger.ear.h <- predict_response(m.bern.finger.ear.h)
plot(pred.m.finger.ear.h, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.finger.ear.h, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.finger.ear.h)

#Running bernouilli model to predict species based on finger length and head
m.bern.finger <- brm(Species ~ 1 + Finger_Length + Head_Length, 
                        data = kratf, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.finger") 

#summary model
summary(m.bern.finger)
plot(m.bern.finger)

#Predict the response based on the model and plot it
pred.m.finger <- predict_response(m.bern.finger)
plot(pred.m.finger, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.finger, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.finger)

```
