---
title: "Generating Graphs EEB symposium"
format: html
editor: visual
---

# 06-01-2026 EEB symposium graph generation

## Set up

I will first load the required packages.

```{r}
library(ggplot2)
library(brms)
library(ggeffects)
library(tidyverse)
library(readxl)
library(here)
library(dplyr)
#install.packages("ggtext")
library(ggtext)
krat <- read.csv(here("final_krat_data_2026.csv"))

```

# Exploring Head vs. Body as size proxy

```{r}
head(krat)
#Plot the body length vs head size
krat %>%
  ggplot(aes(y= Head_Length, x = Body_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
#Plotting standard measurements against eachother
krat %>%
  ggplot(aes(x = Head_Length, y = Foot_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
#Plotting body vs foot
krat %>%
  ggplot(aes(x = Body_Length, y = Foot_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
#plotting head length vs ear length
krat %>%
  ggplot(aes(x = Head_Length, y = Ear_Length_Notch, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
#Plotting body length vs ear length
krat %>%
  ggplot(aes(x = Body_Length, y = Ear_Length_Notch, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")+
   coord_cartesian(ylim = c(10, 30)) #crop out outlier (120 mm in ear length)
#Plotting finger length measurements vs Head length measurements
krat %>%
  ggplot(aes(x = Head_Length, y = Finger_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")


```

# Fixing Ear length outlier

I found an outlier with a value over 120 in the ear length row. Here I fixed that issue using the filter fnct. We can see in these graphs the head length differentiates between the two species better than body length, but that no matter what there is significant overlap in measurements when using the ear length

```{r}
#Filter out the ear outlier
kratf <- krat |> filter(Ear_Length_Notch <30)

#plotting head length vs ear length
kratf %>%
  ggplot(aes(x = Head_Length, y = Ear_Length_Notch, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
#Plotting body length vs ear length
kratf %>%
  ggplot(aes(x = Body_Length, y = Ear_Length_Notch, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

```

# Testing graphs

```{r}
#plotting finger length vs ear length
kratf %>%
  ggplot(aes(x = Finger_Length, y = Ear_Length_Notch, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Running bernouilli model to predict species based on ear, finger, head
m.bern.finger.ear.h <- brm(Species ~ 1 + Ear_Length_Notch + Finger_Length+ Head_Length, 
                        data = kratf, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.body") 

#summary model
summary(m.bern.finger.ear.h)
plot(m.bern.finger.ear.h)

#Predict the response based on the model and plot it
pred.m.finger.ear.h <- predict_response(m.bern.finger.ear.h)
plot(pred.m.finger.ear.h, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.finger.ear.h, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.finger.ear.h)

#Running bernouilli model to predict species based on finger length and head
m.bern.finger <- brm(Species ~ 1 + Finger_Length + Head_Length, 
                        data = kratf, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.finger") 

#summary model
summary(m.bern.finger)
plot(m.bern.finger)

#Predict the response based on the model and plot it
pred.m.finger <- predict_response(m.bern.finger)
plot(pred.m.finger, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.finger, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.finger)

```

# Final visualizations

```{r}
#This is finger length vs. head length
kratf %>%
  ggplot(aes(y= Finger_Length, x = Head_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")+ theme_bw()+
  theme(
    legend.text=element_text(face="italic"),
    legend.position=c(0.85,0.2),
    plot.title=element_markdown(hjust=0.5)
        )+
  scale_color_manual(values=c("darkturquoise","palevioletred"))+
  labs(
    x="Head length",
    y="Finger length",
    title="Finger Length vs. Head length of 
    <br>*Dipodomys heermanni* and *Dipodomys venustus*"
  )
#This is the hand width vs. head length
kratf %>%
  ggplot(aes(y= Hand_Width, x = Head_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")+ theme_bw()+
  theme(
    legend.text=element_text(face="italic"),
    legend.position=c(0.85,0.2),
    plot.title=element_markdown(hjust=0.5)
        )+
  scale_color_manual(values=c("darkturquoise","palevioletred"))+
  labs(
    x="Head length",
    y="Hand Width",
    title="Hand Width vs. Head length of 
    <br>*Dipodomys heermanni* and *Dipodomys venustus*"
  )


```

# Finger length graphs

```{r}
#Running bernouilli model to predict species based on finger length and head
m.bern.finger <- brm(Species ~ 1 + Finger_Length + Head_Length, 
                        data = kratf, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.finger") 

#summary model
summary(m.bern.finger)
plot(m.bern.finger)

#Predict the response based on the model and plot it
pred.m.finger <- predict_response(m.bern.finger)
plot(pred.m.finger, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.finger, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.finger)

#Predict the response based on the model and plot it
pred.m.finger <- predict_response(m.bern.finger)
#plot this
plot(pred.m.finger, show_data = TRUE)
head(pred.m.finger$Finger_Length)
#Make model finger into its own data frame to plot
df.finger <- data.frame(pred.m.finger$Finger_Length)
head(df.finger)


#Create the plot from before but including error and prettier
pred.m.finger <- df.finger|> ggplot(mapping=aes(y=predicted, x=x))+
#Add error bars using the confidence interval as the error bars
geom_errorbar(aes(ymin = conf.low, ymax = conf.high, color = std.error), width = 0.5, linewidth = 0.75)+
  geom_point(size = 2)+ scale_x_continuous(breaks = unique(df.finger$x))+
#Make it pretty
theme_bw()+
  ylim(0,1)+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.5))+
  scale_color_gradient(low= "black", high = "maroon")+
  labs(
    x="Finger length (mm)",
    y="Probability of Species",
    title="Species probability based on finger length and head length values"
  )+
    labs(color = "Error")
print(pred.m.finger)



pred.m.finger <- predict_response(m.bern.finger)
#plot thi
plot(pred.m.finger, show_data = TRUE)
head(pred.m.finger$Head_Length)#a frame to plot
df.head <- data.frame(pred.m.finger$Head_Length)
head(df.head)


#Create the plot from before but including error and prettier
pred.m.finger.head <- df.head |> ggplot(mapping=aes(y=predicted, x=x))+
#Add error bars using the confidence interval as the error bars
geom_errorbar(aes(ymin = conf.low, ymax = conf.high, color = std.error), width = 0.5, linewidth = 0.75)+
  geom_point(size = 2)+ scale_x_continuous(breaks = unique(df.head$x))+
#Make it pretty
theme_bw()+
  #Change y axis to be 0-1
  ylim(0,1)+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.5))+
  scale_color_gradient(low= "black", high = "maroon")+
  labs(
    x="Head length (mm)",
    y="Probability of Species",
    title="Species probability based on finger length and head length values"
  )+
    labs(color = "Error")
print(pred.m.finger.head)


#Create vector using values from model
Pred.sp <- fitted(m.bern.finger)
head(Pred.sp)
#Filter df for only observations including finger length AKA get rid of NAs
#df2 <- df |> filter(Finger_Length )
df_filtered <- kratf[!is.na(kratf$Finger_Length) & !is.na(kratf$Head_Length), ]

nrow(df_filtered)
# Create as column
df_filtered$Probability.sp <- NA
head(df_filtered)

#Create new data frame using the vector
Pred.sp.data <- data.frame(Pred.sp)
head(Pred.sp.data$Estimate)
Pred.sp.est <- Pred.sp.data$Estimate

#showing rows of each
nrow(df_filtered)
length(Pred.sp.data$Estimate)
#Add data estimate to new column
df_filtered$Probability.sp <- Pred.sp.data$Estimate 
head(df_filtered)
#Should we add the error too???


#boxplot!!!!
pred.finger.boxplot <- df_filtered |> ggplot(mapping = aes(y = Probability.sp, x = Species, color = Species))+geom_boxplot()+geom_point()+geom_jitter(width = 0.2, height = 0)+theme_bw()+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.1)
        )+
  scale_color_manual(values=c("darkturquoise","palevioletred"))+
  labs(
    x="Actual Species",
    y="Probability of Species",
    title="Testing Accuracy of finger length and head length based species predictions"
  )
print(pred.finger.boxplot)
```

# Hand width graphs

```{r}
#Running bernouilli model to predict species based on hand W and head L
m.bern.handW <- brm(Species ~ 1 + Hand_Width + Head_Length, 
                        data = kratf, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.handW") 

#summary model
summary(m.bern.handW)
plot(m.bern.handW)

#Predict the response based on the model and plot it
pred.m.handW <- predict_response(m.bern.handW)
plot(pred.m.handW, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.handW, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.handW)

#Predict the response based on the model and plot it
pred.m.handW <- predict_response(m.bern.handW)
#plot this
plot(pred.m.handW, show_data = TRUE)
head(pred.m.handW$Hand_Width)
#Make model finger into its own data frame to plot
df.handW <- data.frame(pred.m.handW$Hand_Width)
head(df.handW)


#Create the plot from before but including error and prettier
pred.m.handW <- df.handW|> ggplot(mapping=aes(y=predicted, x=x))+
#Add error bars using the confidence interval as the error bars
geom_errorbar(aes(ymin = conf.low, ymax = conf.high, color = std.error), width = 0.5, linewidth = 0.75)+
  geom_point(size = 2)+ scale_x_continuous(breaks = unique(df.handW$x))+
#Make it pretty
theme_bw()+
  ylim(0,1)+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.5))+
  scale_color_gradient(low= "black", high = "maroon")+
  labs(
    x="Hand Width (mm)",
    y="Probability of Species",
    title="Species probability based on Hand Width and Head Length Values"
  )+
    labs(color = "Error")
print(pred.m.handW)



pred.m.handW <- predict_response(m.bern.handW)
#plot thi
plot(pred.m.handW, show_data = TRUE)
head(pred.m.handW$Head_Length)#a frame to plot
df.hand.head <- data.frame(pred.m.handW$Head_Length)
head(df.hand.head)


#Create the plot from before but including error and prettier
pred.m.hand.head <- df.hand.head |> ggplot(mapping=aes(y=predicted, x=x))+
#Add error bars using the confidence interval as the error bars
geom_errorbar(aes(ymin = conf.low, ymax = conf.high, color = std.error), width = 0.5, linewidth = 0.75)+
  geom_point(size = 2)+ scale_x_continuous(breaks = unique(df.hand.head$x))+
#Make it pretty
theme_bw()+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.5))+
  scale_color_gradient(low= "black", high = "maroon")+
  labs(
    x="Head length (mm)",
    y="Probability of Species",
    title="Species probability based on hand width and head length values"
  )+
    labs(color = "Error")
print(pred.m.hand.head)


#Create vector using values from model
Pred.sp.h <- fitted(m.bern.handW)
head(Pred.sp.h)

#Filter df for only observations including finger length AKA get rid of NAs
#df2 <- df |> filter(Finger_Length )
df_filtered_h <- kratf[!is.na(kratf$Hand_Width) & !is.na(kratf$Head_Length), ]

nrow(df_filtered_h)
# Create as column
df_filtered_h$Probability.sp <- NA
head(df_filtered_h)

#Create new data frame using the vector
Pred.sp.data.h <- data.frame(Pred.sp.h)
head(Pred.sp.data.h$Estimate)
Pred.sp.est.h <- Pred.sp.data.h$Estimate

#showing rows of each
nrow(df_filtered_h)
length(Pred.sp.data.h$Estimate)
#Add data estimate to new column
df_filtered_h$Probability.sp <- Pred.sp.data.h$Estimate 
head(df_filtered_h)
#Should we add the error too???


#boxplot!!!!
pred.handW.boxplot <- df_filtered_h |> ggplot(mapping = aes(y = Probability.sp, x = Species, color = Species))+geom_boxplot()+geom_point()+geom_jitter(width = 0.2, height = 0)+theme_bw()+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.1)
        )+
  scale_color_manual(values=c("darkturquoise","palevioletred"))+
  labs(
    x="Actual Species",
    y="Predicted Species",
    title="Testing Accuracy of Hand Width and Head Length based species predictions"
  )
print(pred.handW.boxplot)

```

\


#Create new data frame using the vector
Pred.sp.data.h \<- data.frame(Pred.sp)
head(Pred.sp.data.h\$Estimate)
Pred.sp.est.h \<- Pred.sp.data.h\$Estimate

#showing rows of each
nrow(df_filtered)
length(Pred.sp.data\$Estimate)
#Add data estimate to new column
df_filtered\$Probability.sp \<- Pred.sp.data\$Estimate
head(df_filtered)
#Should we add the error too???


#boxplot!!!!
pred.finger.boxplot \<- df_filtered \|\> ggplot(mapping = aes(y = Probability.sp, x = Species, color = Species))+geom_boxplot()+geom_point()+geom_jitter(width = 0.2, height = 0)+theme_bw()+
theme(
legend.text=element_text(face="italic"),
plot.title=element_markdown(hjust=0.5)
)+
scale_color_manual(values=c("darkturquoise","palevioletred"))+
labs(
x="Species",
y="Predictive Power",
title="Testing Accuracy of finger length based species predictions"
)
print(pred.finger.boxplot)
