---
title: "PCA.10.5.2026"
format: html
editor: visual
---

# Info:

I am doing a principle component analysis on the data of Fort Ord and Hastings kangaroo rats for all of my variables. I will be using code from "Introduction to Ordination (PCA) in R" by Brad Duthie and also code from Dr. Rob Burton a CSUMB professor.

I have chosen a PCA over other kinds of ordination because:

-   maximizes total variation across all variables in analysis

-   is blind to groups and would help to hypothesis test for species ID without introducing bias by including grouping by species into analysis

-   Works well with small sample sizes, our is about 100 or less which is relatively small

-   Can help discover unknown data clusters

-   Can create explanations of variation

## Set up

```{r}
library(ggplot2)
library(brms)
library(ggeffects)
library(tidyverse)
library(readxl)
library(here)
library(dplyr)
library(ggtext)

#Load in clean data set
df <- read.csv("RC2026_KRAT_DATA.08.01.csv")
head(df)
```

# Remove NAs set up

```{r}
#Filter out observations where they were recaps or had no data... Use most common measurement, if measurements were taken, body length should have data to filter
df_na <- df |> filter(Body_Length != "NA")
#Show us how many different animals we have left: 67 individuals
print(df_na$Tag)
#Look at data.. how many observations still have NAs???
#print(df_na) 

#Remove non-continuous, non-quantitative variable columns
df_na <- df_na |> select(-OBJECTID, -GlobalID, -Collection_Date, -Tissue_Sample, -Sex, -Reproductive_Stage, -Life_Stage, -Species, -Weight_Bag_Animal, -Genus, -Tag, -YN_Recapture, -Location, -Date, -Latitude, -Longitude, -Notes)

#Find NA count
colSums(is.na(df_na))

#Find rows with highest number of NAs to get rid of
rowSums(is.na(df_na))
#Rows number 12, 23, 40, 43, 48 have A LOT of Nas between 10-13
#Rows number 24, 25,26,27,28 have around 5-6 NAs
# a lot are only missing 2-3 values -- which are bolea and foot width consider deleting these columns

#remove rows rows with most NAs
df_na1 <- df_na[-c(12, 23, 40, 43, 48, 24, 25, 26, 27, 28), ]
df_na1 <- df_na1 |> select(-Bolea_Head_Width, -Foot_Width)
#Check how many NAs we still have
colSums(is.na(df_na1))
rowSums(is.na(df_na1))
#This way we have way less NAs
#Only columns to look into now are animal weight and forelimb length for having missing data
#print(df_na1)

#all NAs omitted leaves us with only 25 observations :C
df_na_omit <- na.omit(df_na)
#print(df_na_omit)
```

Looking at the data after filtering out the observations where we likely do not have any measurements, we can see we still have a significant amount of data with a lot of NAs...

Consider not including measurements that have too many NAs??? such as bolea head width and foot width

For others such as foot length we could consider inputting the mean/median?

## Set up Matrix

```{r}
#Show column names
colnames(df)

#Remove non-continuous, non-quantitative variable columns
df <- df |> select(-OBJECTID, -GlobalID, -Collection_Date, -Tissue_Sample, -Sex, -Reproductive_Stage, -Life_Stage, -Species, -Weight_Bag_Animal, -Genus, -Tag, -YN_Recapture, -Location, -Date, -Latitude, -Longitude, -Notes)
#Check that everything is right
colnames(df)
#turn it into a matrix for the PCA
df <- as.matrix(df)
#This matrix exists in 15 dimensional space
head(df)
```

# Visualize data

```{r}
#show data in pairs
pairs(x = df, gap = 0, cex.labels = 0.5)
#SCARYYYYYY
```

```{r}
#Scale data so that all data has a mean of 0 and SD of 1
df <- scale(df)
head(df)
#try to make the pca
pca_df <- prcomp(df)


```

It does not work because I have too many NAs in the dataset, if I omit them I would lose too much data

My options are:

-   fill in missing value NAs with a mean or median

-   regularized iterative PCA

-   NIPALS – Non-linear Iterative Partial Least Squares
