---
title: "PCA.10.5.2026"
format: html
editor: visual
---

# Info:

I am doing a principle component analysis on the data of Fort Ord and Hastings kangaroo rats for all of my variables. I will be using code from "Introduction to Ordination (PCA) in R" by Brad Duthie and also code from Dr. Rob Burton a CSUMB professor.

I have chosen a PCA over other kinds of ordination because:

-   maximizes total variation across all variables in analysis

-   is blind to groups and would help to hypothesis test for species ID without introducing bias by including grouping by species into analysis

-   Works well with small sample sizes, our is about 100 or less which is relatively small

-   Can help discover unknown data clusters

-   Can create explanations of variation

## Set up

```{r}
library(ggplot2)
library(brms)
library(ggeffects)
library(tidyverse)
library(readxl)
library(here)
library(dplyr)
library(ggtext)

#Load in clean data set
df <- read.csv("RC2026_KRAT_DATA.08.01.csv")
head(df)
```

## Set up Matrix

```{r}
#Show column names
colnames(df)

#Remove non-continuous, non-quantitative variable columns
df <- df |> select(-OBJECTID, -GlobalID, -Collection_Date, -Tissue_Sample, -Sex, -Reproductive_Stage, -Life_Stage, -Species, -Weight_Bag_Animal, -Genus, -Tag, -YN_Recapture, -Location, -Date, -Latitude, -Longitude, -Notes)
#Check that everything is right
colnames(df)
#turn it into a matrix for the PCA
df <- as.matrix(df)
#This matrix exists in 15 dimensional space
head(df)
```

# Visualize data

```{r}
#show data in pairs
pairs(x = df, gap = 0, cex.labels = 0.5)
#SCARYYYYYY
```

```{r}
#Scale data so that all data has a mean of 0 and SD of 1
df <- scale(df)
head(df)
#try to make the pca
pca_df <- prcomp(df)


```

It does not work because I have too many NAs in the dataset, if I omit them I would lose too much data

My options are:

-   fill in missing value NAs with a mean or median

-   regularized iterative PCA

-   NIPALS – Non-linear Iterative Partial Least Squares
