---
title: "PCA.10.5.2026"
format: html
editor: visual
---

# Info:

I am doing a principle component analysis on the data of Fort Ord and Hastings kangaroo rats for all of my variables. I will be using code from "Introduction to Ordination (PCA) in R" by Brad Duthie and also code from Dr. Rob Burton a CSUMB professor.

I have chosen a PCA over other kinds of ordination because:

-   maximizes total variation across all variables in analysis

-   is blind to groups and would help to hypothesis test for species ID without introducing bias by including grouping by species into analysis

-   Works well with small sample sizes, our is about 100 or less which is relatively small

-   Can help discover unknown data clusters

-   Can create explanations of variation

## Set up

```{r}
library(ggplot2)
library(brms)
library(ggeffects)
library(tidyverse)
library(readxl)
library(here)
library(dplyr)
library(ggtext)

#Load in clean data set
df <- read.csv("RC2026_KRAT_DATA.08.01.csv")
head(df)
```

# Sort out NAs try 1

```{r}

#Remove non-continuous, non-quantitative variable columns
df <- df |> select(-OBJECTID, -GlobalID, -Collection_Date, -Tissue_Sample, -Sex, -Reproductive_Stage, -Life_Stage, -Species, -Genus, -Tag, -YN_Recapture, -Location, -Date, -Latitude, -Longitude, -Notes)

#Show NA values in full clean data set
print(df)
#Show NAs in columns and rows
colSums(is.na(df))
rowSums(is.na(df))

#Remove rows with the most NAs
df <- df[-c(6, 10, 11, 12,13, 14,15, 16, 19, 20, 21, 22, 23, 24, 25, 34, 36, 37, 38, 39, 51, 57, 60, 61, 66, 69, 70, 76, 78, 79, 80, 86, 90, 91, 92, 93, 95, 96), ]
colSums(is.na(df))
rowSums(is.na(df))
#This cleaning leaves 60 observations
print(df)

#Some Weight_animal_only have Weight_bag_animal values which can be estimates to find the weight or subtracted the average of bag weight (Usually between 10 and 20 based on how many seeds) -- if we do that there are only 3/8 truly missing values

df.test$Na_Animal_Weight <- df$Weight_Bag_Animal-15
head(df.test)
df.test$Weight_Animal_Only[is.na(df.test$Weight_Animal_Only)] <- df.test$Na_Animal_Weight[is.na(df$Weight_Animal_Only)]
#It worked!!! now only 3 missing weight values
colSums(is.na(df.test))
#Get rid of unwanted columns
df.test <- df.test |> select(-Weight_Bag_Animal,-Na_Animal_Weight)

#Will get rid of Weight_Bag_animal once animal  only values are cleaned up

#Potentially get rid of Bolea_head_width and Foot_width from PCA analysis, they have 18 and 17 missing values

```

# Remove NAs set up

```{r}
#Filter out observations where they were recaps or had no data... Use most common measurement, if measurements were taken, body length should have data to filter
df_na <- df |> filter(Body_Length != "NA")
#Show us how many different animals we have left: 67 individuals
print(df_na$Tag)
#Look at data.. how many observations still have NAs???
#print(df_na) 

#Remove non-continuous, non-quantitative variable columns
df_na <- df_na |> select(-OBJECTID, -GlobalID, -Collection_Date, -Tissue_Sample, -Sex, -Reproductive_Stage, -Life_Stage, -Species, -Weight_Bag_Animal, -Genus, -Tag, -YN_Recapture, -Location, -Date, -Latitude, -Longitude, -Notes)

#Find NA count: 139 total out of
colSums(is.na(df_na))

#Find rows with highest number of NAs to get rid of
rowSums(is.na(df_na))
#Rows number 12, 23, 40, 43, 48 have A LOT of Nas between 10-13
#Rows number 24, 25,26,27,28 have around 5-6 NAs
# a lot are only missing 2-3 values -- which are bolea and foot width consider deleting these columns

#remove rows rows with most NAs
df_na1 <- df_na[-c(12, 23, 40, 43, 48, 24, 25, 26, 27, 28), ]
df_na1 <- df_na1 |> select(-Bolea_Head_Width, -Foot_Width)
#Check how many NAs we still have, 25 in total over 57*13 rows*columns
#Thats 3.37 % of values are missing -- shouldnt be too bad???
colSums(is.na(df_na1))
rowSums(is.na(df_na1))
#This way we have way less NAs
#Only columns to look into now are animal weight and forelimb length for having missing data
print(df_na1)

#all NAs omitted leaves us with only 25 observations :C
df_na_omit <- na.omit(df_na)
print(df_na_omit)
```

Looking at the data after filtering out the observations where we likely do not have any measurements, we can see we still have a significant amount of data with a lot of NAs...

Consider not including measurements that have too many NAs??? such as bolea head width and foot width

For others such as foot length and ones with only 1 missing data piece we could consider inputting the mean/median?

For animal weight consider doing an estimate of bag weight if bag + animal value exists

Forelimb length has about 7 values missing. – what to do?

## Set up Matrix

```{r}
#Show column names
colnames(df)

#Remove non-continuous, non-quantitative variable columns
df <- df |> select(-OBJECTID, -GlobalID, -Collection_Date, -Tissue_Sample, -Sex, -Reproductive_Stage, -Life_Stage, -Species, -Weight_Bag_Animal, -Genus, -Tag, -YN_Recapture, -Location, -Date, -Latitude, -Longitude, -Notes)
#Check that everything is right
colnames(df)
#turn it into a matrix for the PCA
df <- as.matrix(df)
#This matrix exists in 15 dimensional space
head(df)
```

# Visualize data

```{r}
#show data in pairs
pairs(x = df, gap = 0, cex.labels = 0.5)
#SCARYYYYYY
```

```{r}
#Scale data so that all data has a mean of 0 and SD of 1
df <- scale(df)
head(df)
#try to make the pca
pca_df <- prcomp(df)


```

It does not work because I have too many NAs in the dataset, if I omit them I would lose too much data

My options are:

-   fill in missing value NAs with a mean or median

-   regularized iterative PCA

-   NIPALS – Non-linear Iterative Partial Least Squares
