---
title: "Raw to Clean data"
format: html
editor: visual
---

This code steps 1-5 was written as part of an assignment for Data Science for EEB class in 2026 using this same data set by Yeonwoo Lee and Mackenzie Black. And I (Rowan Crowley) will now be including and cleaning more recent data taken between January 29, 2026 and July 26, 2026.

## Set up + Load data

```{r}
library(ggplot2)
library(brms)
library(ggeffects)
library(tidyverse)
library(readxl)
library(here)
library(dplyr)
#install.packages("ggtext")
library(ggtext)
#Load in sorted data from 1/29
#df1 <- read_excel("1_29Kangaroo Rat Data_TableToExcel.xlsx")
#df2 is oldest data before november 2025 before krat n other data was split
df2 <- read_excel("1_29Small_Mammal_Observations_TableToExcel.xlsx")
#Load in new data most recent 7_26 -- from this summer after june 30,2026
df3 <- read_excel("New_App_krat_data_7_31_2026.xls") 
#7_26 oldkratdata2026 is the same dataset as 1_29krat data but with new values added... replace any df1
df1<- read_excel("7_31oldkratdata_2026.xls")

```

# Step 1: Fix df2, oldest data 2025

### Filter Dipodomys df2

```{r}
df2_Difiltered <- df2 %>%
  #filter for DI, Di, dI, di
  filter(tolower(genus) == "di")

```

### Extract Head widths df2

```{r}
#Distinguish pm + b
df2_head_mb_cleaned <- df2_Difiltered %>%
  mutate(
    # Extract all numeric values and labels (pm, b) in their appearing order
    all_nums = str_extract_all(KRAT_Head_Width_mm, "[0-9.]+"),
    all_labels = str_extract_all(str_to_lower(KRAT_Head_Width_mm), "pm|b"),
    
    # Map the first and second extracted elements to temporary variables
    num1 = sapply(all_nums, function(x) as.numeric(x[1])),
    num2 = sapply(all_nums, function(x) as.numeric(x[2])),
    lab1 = sapply(all_labels, function(x) x[1]),
    lab2 = sapply(all_labels, function(x) x[2])
  ) %>%
  mutate(
    # Rule: Assign num1/num2 to Maxillary or Bolea based on the order of lab1/lab2
    # If no label exists, num1 defaults to Maxillary_Head_Width
    Maxillary_Head_Width = case_when(
      is.na(lab1) ~ num1,
      lab1 == "pm" ~ num1,
      lab2 == "pm" ~ num2,
      TRUE ~ NA_real_
    ),
    Bolea_Head_Width = case_when(
      lab1 == "b" ~ num1,
      lab2 == "b" ~ num2,
      TRUE ~ NA_real_
    )
  )
print(df2_head_mb_cleaned)
#Extract forelimb length???? from notes
```

### Distribute notes data into their own column df2

#move tissue sample values, forelimb length values, KRAT_Foot_Length values to foot length, bolea, maxillary, tissue samples from notes, weight calculations done, location name corrected, etc

```{r}
#Create a csv so I can edt in excel
#write_csv(df2_head_mb_cleaned, "krat_hand_edited.....csv")

df2_edited <- read.csv("krat_hand_edited_2025.csv") 
colnames(df2_edited)

```

# Step 2: Fix df1, old app 2026, no more data to be added

### Find lat/long for df1

Tried trouble shooting on arcgis pro but unfortunately I do not have the permission to do this on my GIS account. I will need to ask Joe to get a csv that includes the lat/long of the GIS online app titled "Mammal_Observation_Database_Fort_Ord_Natural_Reserve" which is not currently included in the attribute table, yet we have the points on the maps... weirddd

```{r}
#dont have permission to do it, will need to ask joe!!!
```

### Fix tissue sample format df1

```{r}
#Now editing tissue samples to be formatted: "RC2025-##"
df1$Tissue_Sample <- paste0("RC2025-", df1$Tissue_Sample)
#Messed up now double RC2025 in 3 rows... :( will need to fix later
#print(df1$Tissue_Sample)
```

# Step 3: rename all df for merge

### Standard names are:

OBJECTID

GlobalID

Latitude

Longitude

Notes

Tissue_Sample

Hind_Limb_Length

Foot_Width

Wrist_Width

Forelimb_Length

Bolea_Head_Width

Maxillary_Head_Width

Head_Length

Teeth

Finger_Length

Hand_Width

Ear_Length

Toe_Length

Foot_Length

Tail_Length

Body_Length

Reproductive_Stage

Sex

Life_Stage

Weight_Animal_Only

Weight_Bag_Animal

Species

Genus

Tag

YN_Recapture

YN_Animal

Location

Date

```{r}
#df1 renamed for merge

df1_renamed <- df1 |> rename(
    Forelimb_Length = Forelimb_Length_Nail_to_Elbow,
    Ear_Length = Ear_Length_Notch,
    Location = LOC,
    Date = DT) |> select( -EditDate, -Editor, -Teeth)

#df2 renamed for merge
df2_renamed <- df2_edited |>
  rename(
OBJECTID = OBJECTID,
GlobalID = GlobalID,
CreationDate = CreationDate,
Notes = KRAT_Notes,
Forelimb_Length = KRAT_Forelimb_Length_mm,
Hind_Limb_Length = KRAT_Hind_Limb_Length,
Foot_Width = KRAT_Foot_Width_mm,
Wrist_Width = KRAT_Wrist_Width_mm,
Head_Length = KRAT_Head_Length_mm,
Finger_Length = KRAT_Finger_Length,
Hand_Width = KRAT_Hand_Width_mm,
Ear_Length = Ear_mm,
Toe_Length = KRAT_Hind_Toe_Length,
Foot_Length = Foot_mm,
Tail_Length = Tail_mm,
Body_Length = Body_mm,
Reproductive_Stage = Reproductive_Stage,
Sex = Sex,
Life_Stage = Life_Stage,
Weight_Animal_Only = Animal_Weight_g,
Species = Species_Code,
Genus = genus,
Tag = Tag_Number,
YN_Recapture = Recap_y_n,
YN_Animal = Animal_y_n,
Location = location_code,
Date = date_time,
Latitude = esrignss_latitude,
Longitude = esrignss_longitude) |> 
  #Remove irrelevant columns
  select(-esrignss_speed, -esrignss_direction, -esrisnsr_azimuth, -esrignss_positionsourcetype, -esrignss_receiver, -esrignss_h_rms, -esrignss_v_rms, -esrignss_altitude, -esrignss_pdop, -esrignss_hdop, -esrignss_vdop, -esrignss_fixtype, -esrignss_correctionage, -esrignss_stationid, -esrignss_numsats, -esrignss_fixdatetime, -esrignss_avg_h_rms, -esrignss_avg_positions, -esrignss_h_stddev, -EditDate, -Editor, -Subspecies_Code, -Tripped_Trap_y_n, -Bait_Stolen_y_n, -KRAT_Teeth, -KRAT_Head_Width_mm, -esrignss_avg_v_rms, -all_nums, -all_labels, -num1, -num2, -lab1, -lab2, -KRAT_Foot_Length_mm)

head(df2_renamed)

#df3 renamed for merge
df3_renamed <- df3 |>
rename(OBJECTID = OBJECTID,
       GlobalID = GlobalID,
       CreationDate = CreationDate,
       Notes = Notes,
       Tissue_Sample = Tissue_Sample,
       Hind_Limb_Length = Hind_limb_length,
       Foot_Width =Foot_Width,
       Wrist_Width = Wrist_Width,
       Forelimb_Length = forelimb_L_nail_to_elbow,
       Bolea_Head_Width = Bolea_Head_Width,
       Maxillary_Head_Width = Maxillary_Head_Width,
       Head_Length = Head_L,
       Finger_Length = Finger_L,
       Hand_Width = Hand_w,
       Ear_Length = Ear_L,
       Toe_Length = Toe_L,
       Foot_Length = Foot_L,
       Tail_Length = Tail_L,
       Body_Length = Body_L,
       Reproductive_Stage = Reproductive_Stage,
       Sex = Sex,
       Life_Stage = Life_Stage,
       Weight_Animal_Only = Weight_animal_only,
       Weight_Bag_Animal = Weight_baganimal,
       Species = Species,
       Genus = Genus,
       Tag = tag_number,
       YN_Recapture = YN_Recapture,
       YN_Animal =YN_Animal,
       Location = Location_name,
       Date = Date_and_time,
       Latitude = esrignss_latitude,
       Longitude = esrignss_longitude) |>
  #Remove ireelevent columns
select( -esrignss_receiver, -esrignss_positionsourcetype, -esrignss_altitude, -esrignss_orthometric_height, -esrignss_h_rms, -esrignss_v_rms, -esrignss_fixdatetime, -esrignss_fixtype, -esrignss_numsats, -esrignss_correction_source_type, -esrignss_correction_source_name, -esrignss_correctionage, -esrignss_ntrip_mount_point, -esrignss_stationid, -esrignss_pdop, -esrignss_hdop, -esrignss_vdop, -esrignss_direction, -esrignss_speed, -esrisnsr_azimuth, -esrignss_antenna_height, -esrignss_receiver_gcs, -esrignss_receiver_gcs_wkid, -esrignss_receiver_ref_epoch, -esrignss_profile_has_hor_xform, -esrignss_profile_hor_xform, -esrignss_receiver_geoid, -esrignss_avg_h_rms, -esrignss_avg_v_rms, -esrignss_avg_positions, -esrignss_h_stddev, -esrignss_h_95_status, -esrignss_custom_attributes, -Teeth, -EditDate, -Editor)

colnames(df3)
```

# Step 4: As numeric, all df etc

```{r}
# Data type change to numeric
colnames(df1_renamed)
df1_numeric <- df1_renamed %>%
  mutate(across(c(Bolea_Head_Width, Maxillary_Head_Width, Toe_Length, Hind_Limb_Length, Foot_Width, Wrist_Width, Head_Length, Finger_Length, Forelimb_Length, Ear_Length, Tail_Length, Body_Length, Hand_Width, Foot_Length, Tail_Length, Weight_Bag_Animal, Weight_Animal_Only), as.numeric))
head(df1_numeric)


df2_numeric <- df2_renamed %>%
  mutate(across(c(Bolea_Head_Width, Maxillary_Head_Width, Toe_Length, Hind_Limb_Length, Foot_Width, Wrist_Width, Head_Length, Finger_Length, Forelimb_Length, Ear_Length, Tail_Length, Body_Length, Hand_Width, Foot_Length, Tail_Length, Weight_Animal_Only), as.numeric))
#Make Creation date for df2 from <character> to <datetime<UTC>>
df2_numeric2.0 <- df2_numeric |> mutate(CreationDate = ymd_hms(CreationDate)) |> mutate(Date = ymd_hms(Date))
head(df2_numeric2.0)
#Working on df3
df3_numeric <- df3_renamed |>  mutate(across(c(Bolea_Head_Width, Maxillary_Head_Width, Toe_Length, Hind_Limb_Length, Foot_Width, Wrist_Width, Head_Length, Finger_Length, Forelimb_Length, Ear_Length, Tail_Length, Body_Length, Hand_Width, Foot_Length, Tail_Length, Weight_Animal_Only, Weight_Bag_Animal, Weight_bag_only), as.numeric))
```

# 

# Step 4.5: Bag weight calc df3

---ALSO edited bag weight to subtract for animal weight in df3

```{r}
df3_numeric <- df3_numeric |> mutate(Weight_Animal_Only = Weight_Bag_Animal - Weight_bag_only)
print(df3_numeric$Weight_Animal_Only)
print(df3_numeric$Weight_bag_only)
```

# Step 5: MERGE all df

```{r}
# Bind them together
krat_data1 <- bind_rows(df1_numeric, df2_numeric2.0)
head(krat_data1)
krat_data2 <- bind_rows(krat_data1, df3_numeric)
```

# Step 6: species assignment & type conversion

```{r}
final_krat_data <- krat_data2 %>%
  # 1. Assign Species name based on Location (LOC) column
  mutate(Species = case_when(
    Location == "Hastings" ~ "Dipodomys venustus",
    Location == "FONR"     ~ "Dipodomys heermanni", 
    TRUE ~ Species
  )) %>%
  # 2. Ensure all measurement columns are numeric to prevent plotting errors
  mutate(across(c(contains("Length"), contains("Width"), contains("Weight")), as.numeric))




```

# Step 7: Iron out fine details

### Merge notes columns together, get rid of unimportant columns

```{r}
#Combine both notes columns into one using unite() --- BADDD dont use
#final_krat_data$Notes <- unite(final_krat_data,Notes, NOTES, sep = " ")
final_krat_data$notes <- paste(final_krat_data$Notes, final_krat_data$NOTES)
#Checking that this worked!!! CHECK.
#print(final_krat_data$notes)
#print(final_krat_data$Notes)
#print(final_krat_data$NOTES)

#Get rid of unused Notes and NOTES columns and bag weight column
final_krat_data <- final_krat_data |> select( -NOTES, -Notes, -Weight_bag_only, -YN_Animal)

#Now editing tissue samples to be formatted: "RC2025-##"
#final_krat_data$Tissue_Sample <- paste0("RC2025-", final_krat_data$Tissue_Sample)
#Messed up now double RC2025 in some rows... :(
#print(final_krat_data$Tissue_Sample)
#Fix in excel???? -- dont know if this code is necessary
#print(df1)
```

# Step 8: Fix NAs in data??? (INCOMPLETE)

For columns:

Lifestage – if NA assume Adult

Reproductive stage – if NA assume non-repro

```{r}
#final_krat_data <- final_krat_data %>%
  # 1. Assign default stages for data with NAs
  #mutate(Reproductive_Stage = case_when(
   # Reproductive_Stage == "NA" ~ "Non-Reproductive",
   # TRUE ~ Reproductive_Stage
 # ))
#print(final_krat_data$Reproductive_Stage)
#dont work :(
```

# Step 9: Export csv... ALL CLEAN :)

```{r}
# install.packages("readr")
library(readr)
write_csv(final_krat_data, "Clean...kratdata7_31_2026.csv")
#Manually fix tissue samples in excel
FINALdf <- read.csv("Clean...kratdata7_31_2026.CSV")
unique(FINALdf$Tissue_Sample)
#Delete columns: creator
#rename Creation date to Collection Date, notes to Notes
FINALdf <- FINALdf |> select( -Creator) |> rename( Collection_Date = CreationDate,
Notes = notes)
#show edited df.... ITS GORGEOUS!!!
head(FINALdf)
write_csv(FINALdf, "RC2026_KRAT_DATA.08.01.csv")
#only thing we would need to do is recover the latitude n longitude to make it complete
```

# SIDEWORK: Getting body measurements for dominique

```{r}
dominiquekratdata <- final_krat_data |> select (Body_Length)
dominiquekratdata$Body_Length |> mean(na.rm = TRUE)
#dihe body length mean: 105.5821 mm
#print(dominiquekratdata$Body_Length)
#Lowest value recorded 92 mm, highest value recorted 128 mm
```
