---
title: "Raw to Clean data"
format: html
editor: visual
---

This code steps 1-5 was written as part of an assignment for Data Science for EEB class in 2026 using this same data set by Yeonwoo Lee and Mackenzie Black. And I (Rowan Crowley) will now be including and cleaning more recent data taken between January 29, 2026 and July 26, 2026.

## Set up + Load data

```{r}
library(ggplot2)
library(brms)
library(ggeffects)
library(tidyverse)
library(readxl)
library(here)
library(dplyr)
#install.packages("ggtext")
library(ggtext)
#Load in sorted data from 1/29
#df1 <- read_excel("1_29Kangaroo Rat Data_TableToExcel.xlsx")
#df2 is oldest data before november 2025 before krat n other data was split
df2 <- read_excel("1_29Small_Mammal_Observations_TableToExcel.xlsx")
#Load in new data most recent 7_26 -- from this summer after june 30,2026
df3 <- read_excel("Krat_data_7.26.26.xlsx") 
#7_26 oldkratdata2026 is the same dataset as 1_29krat data but with new values added... replace any df1
df1<- read_excel("7_26oldkratdata2026.xls")
```

# Step 1: Filter Dipodomys

```{r}
df2_Difiltered <- df2 %>%
  #filter for DI, Di, dI, di
  filter(tolower(genus) == "di")

```

# Step 2: Extract Head widths

```{r}
#Distinguish pm + b
df2_head_mb_cleaned <- df2_Difiltered %>%
  mutate(
    # Extract all numeric values and labels (pm, b) in their appearing order
    all_nums = str_extract_all(KRAT_Head_Width_mm, "[0-9.]+"),
    all_labels = str_extract_all(str_to_lower(KRAT_Head_Width_mm), "pm|b"),
    
    # Map the first and second extracted elements to temporary variables
    num1 = sapply(all_nums, function(x) as.numeric(x[1])),
    num2 = sapply(all_nums, function(x) as.numeric(x[2])),
    lab1 = sapply(all_labels, function(x) x[1]),
    lab2 = sapply(all_labels, function(x) x[2])
  ) %>%
  mutate(
    # Rule: Assign num1/num2 to Maxillary or Bolea based on the order of lab1/lab2
    # If no label exists, num1 defaults to Maxillary_Head_Width
    Maxillary_Head_Width = case_when(
      is.na(lab1) ~ num1,
      lab1 == "pm" ~ num1,
      lab2 == "pm" ~ num2,
      TRUE ~ NA_real_
    ),
    Bolea_Head_Width = case_when(
      lab1 == "b" ~ num1,
      lab2 == "b" ~ num2,
      TRUE ~ NA_real_
    )
  )

```

# Step 3: rename for merge

```{r}
df2_renametodf1 <- df2_head_mb_cleaned %>%
  rename(
    Notes = KRAT_Notes,
    Toe_Length = KRAT_Hind_Toe_Length,
    Hind_Limb_Length = KRAT_Hind_Limb_Length,
    Foot_Width = KRAT_Foot_Width_mm,
    Foot_Length = KRAT_Foot_Length_mm,
    Hand_Width = KRAT_Hand_Width_mm,
    Finger_Length = KRAT_Finger_Length,
    Wrist_Width = KRAT_Wrist_Width_mm,
    Head_Length = KRAT_Head_Length_mm,
    Tag = Tag_Number,
    Weight_Animal_Only = Animal_Weight_g,
    LOC = location_code,
    Ear_Length_Notch = Ear_mm,
    Tail_Length = Tail_mm,
    Body_Length = Body_mm,
    Genus = genus
  ) %>%
  # Remove irrelevant or temporary columns (teeth: No records, Head width-: changed to other columns)
  select(-KRAT_Teeth, -KRAT_Head_Width_mm, -all_nums, -all_labels, -num1, -num2, -lab1, -lab2)

#Rename df3 for merge
head(df3)
df3_renamed <- df3 |>
    rename(
    Toe_Length = Toe_L,
    Hind_Limb_Length = Hind_limb_length,
    Forelimb_Length_Nail_to_Elbow = forelimb_L_nail_to_elbow,
    Foot_Length = Foot_L,
    Hand_Width = Hand_w,
    Finger_Length = Finger_L,
    Head_Length = Head_L,
    Tag = tag_number,
    Weight_Animal_Only = Weight_animal_only,
    Weight_Bag_Animal = Weight_baganimal,
    LOC = Location_name,
    Ear_Length_Notch = Ear_L,
    Tail_Length = Tail_L,
    Body_Length = Body_L)
head(df3_renamed)
```

# Step 4: Merge df1 + df2

```{r}
# Data type change to numeric
df1_numeric <- df1 %>%
  select(-Teeth) %>%
  mutate(across(c(Bolea_Head_Width, Maxillary_Head_Width, Toe_Length, Hind_Limb_Length), as.numeric))

df2_numeric <- df2_renametodf1 %>%
  mutate(across(c(Bolea_Head_Width, Maxillary_Head_Width, Toe_Length, Hind_Limb_Length), as.numeric))

#Working on df3
df3_numeric <- df3_renamed |>  mutate(across(c(Bolea_Head_Width, Maxillary_Head_Width, Toe_Length, Hind_Limb_Length), as.numeric))

# Bind them together
krat_data1 <- bind_rows(df1_numeric, df2_numeric)
head(krat_data1)
krat_data2 <- bind_rows(krat_data1, df3_numeric)
write_csv(krat_data2, "krat_data_2.1_2026.csv")
```

# Step 5: species assignment & type conversion

```{r}
final_krat_data <- krat_data2 %>%
  # 1. Assign Species name based on Location (LOC) column
  mutate(Species = case_when(
    LOC == "Hastings" ~ "Dipodomys venustus",
    LOC == "FONR"     ~ "Dipodomys heermanni", 
    TRUE ~ Species
  )) %>%
  # 2. Ensure all measurement columns are numeric to prevent plotting errors
  mutate(across(c(contains("Length"), contains("Width"), contains("Weight")), as.numeric))

# install.packages("readr")
library(readr)
write_csv(final_krat_data, "final_krat_data_2026.csv")

```
